\chapter{PAWS - Towards PIPS as Web Service}

This first chapter introduces the idea of creating PaWS framework. It
answers two questions: why it is needed? and on what it is based? Design
constraints and functional requirements also are presented here. They
resulted in the emergence of three main modes for PaWS. They are
described in the next section (?). The last part of this chapter presents
the creation process of PaWS.

\input{reportCh1/paws}
\input{reportCh1/constraints}
\input{reportCh1/requirements}
\input{reportCh1/goals}
\input{reportCh1/process}

\section{Conclusion}

PaWS is a lightweight and easy to use application, although it refers to a very complex toolkit, which is PIPS. Design contraints and requirements which were evolving over the time and also different modes of using this framework with significant amount of examples resulted in necessity of providing flexibility in configuring and changing PaWS. It required using good architectural solution and proven tools or libraries. These issues are described in subsequent chapters.

``Web service'' in the PaWS acronim should not be taken literally. It is not the implementation of the Web Service standard \cite{ws_standard}. PaWS framework idea refers to the conception of the service\footnote{Service is the independent software element which shares complete, reusable  functionality through the well defined interface \cite{service_def}.} in general. The term ``web'' in this case means that functionality shared by the PaWS service is available via the Internet.
