\section{Design requirements}
\label{design_requirements}

PaWS framework has several functional requirements:

\begin{enumerate}
  \item {\bf Uploading user files (also multiple files)}\label{req:uploading_files}: user can provide the source code not only by typing or copy-and-pasting it into the application window, but also by browsing and picking it from his/her machine.
  \item {\bf Saving results on user machine}\label{req:saving_results}: user can write back the result of the analysis or transformation as a file on his/her machine.
  \item {\bf Printing results}\label{req:printing_results}: user can print the result of the analysis or transormation on his/her printer.
  \item {\bf Presenting dependence graphs}\label{req:dependence_graphs}: user can create and display dependence graph of his/her source code. Also demonstrations can include graphs.
  \item {\bf Detecting language}\label{req:language_detection}: programming language of the source code is detected and displayed. C, Fortran77 and Fortran 95 are supported.
  \item {\bf Deleting temporary files}\label{req:deleting_files}: temporary files, which are not used anymore, are removed from the server disk space.
  \item {\bf Providing ready examples}\label{req:providing_examples}: each analysis and transformation has the set of pedagogical examples linked to it and ready to use for the user.
  \item {\bf Authentication mechanism}\label{req:authentication}: access to the PaWS is protected by a simple authentication mechanism.
\end{enumerate}