\section{Encountered problems}
\label{encountered_problems}

PaWS implementation process encountered several problems and bugs:

\begin{itemize}
  \item {\bf Perserving consistency between output of Tpips and Pyps computations}
  
  At the beginning operations over prepared examples where performed with Tpips script, while user provided code was handled by Pyps. Also supplied example after user modification were analyze or transformed with Pyps. That was causing lack of consistency in the output - Pyps in the basic level is working with the default values, while in the Tpips script other settings can be used as well. Also style of the output performed by the Tpips and Pyps is slightly different. To provide completely consistent output all operation in the tools mode (for basic and for advanced levels) are performed by Pyps. Tpips is used in demonstration mode (which needs script with a set of well-defined operations, which is not modified by a user).
  
  \item {\bf Validation of the Pyps code}
  
  PIPS and Pyps are constantly in the development process. That can cause that the Pyps code used for the PaWS operations might require changes according to the Pyps changes. To minimize problems resulting from this, such as getting a wrong output or being unable to perform operations, all of the Pyps code used in PaWS is moved to the \emph{validation} directory, where it can undergo validation process and be easily changed if it is needed.
  
  \item {\bf Updating PIPS automatically}
  
  This problem is related to the previous one - PIPS framework, which is still changing, has to be update frequently, to let the users try the newest version of it. With automatic update, there might be a problem when current version of PIPS contains bugs or is even broken. To avoid such situation PIPS update is made manually.
  
  \item {\bf Provide concurrent access}
  
  Pylons framework creates new thread for each user request. It might cause problems with Pyps which requires new process. Pyrops (see Section \ref{other_technologies}) module, which separates Pyps workspaces, was the solution of this problem. Pyrops is also under development (it is build in Pyps).
  %% bug with sharing workspaces?
  
  \item {\bf Overloading server}
  
  Too many user requests might overload the server. Problem was not solved for the time being.
  
  \item {\bf Points to the documentation}
  
  User should have information about the PIPS properties and analyses which might be used in advanced level of tools mode. The best option to provide it is by linking them to the proper sections in the PIPS online documentation (\url{http://cri.ensmp.fr/pips/pipsmake-rc.htdoc/}). Problem is with extracting exact addresses of the sections - they are generated randomly and changing during the documentation compilation. Currently information about the PIPS settings (properties, analyses and phases) is provided by the description files.

\end{itemize}